```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```
# Automated Reporting {#auto-report}

Effective communication of results is among the essential duties of the sensory scientist, but the sometimes tedious mechanics of report production together with the sheer volume of data that many scientists now must process combine to make reporting design an afterthought in too many cases.  In this tutorial, we review recent advances in automated report production that liberate resources for scientists to focus on the interpretation and communication of results, while simultaneously reducing errors and increasing the consistency of their analyses. We teach the tutorial through an extended example, cumulatively building an R script that takes participates from receipt of an example dataset to a beautifully-designed and nearly completed PowerPoint presentation automatically and using freely available, open-source packages.  Details of how to customize the final presentation to incorporate corporate branding - such as logos, font choices, and color palettes - will also be covered.

## What is Automated Reporting?
Why Script?
Save time
Reduce errors
Collaboration
Share code with others
Read own code later
Explain choices for analysis, table presentation, charts
Save steps for result creation
Main tools
R/RStudio
RMarkdown, HTML output, etc. (mention but don’t focus)
Packages for Microsoft office production
Officer suite (PowerPoint, Word)
Charts, especially RVG
Extract from Word/PowerPoint
Index
Flextable
Images?
Packages for formatting
extrafont
extrafontdb
Rcolorbrewer

## Excel

Although Excel is not our preferred tool for automated reporting, it is still one of the major ways to access and share data. Most data collection software offer the possibility to export data and results in an Excel format, and most data analysis tools accept Excel format as inputs.  With the large use of Excel, it is no surprise that many of our colleagues or clients like to share data and results using such spreadsheets. It is even less a surprise that R provides multiple solutions to import/export results from/to Excel. 

For the import of datasets, we have already presented the package `{readxl}` among others. For exporting results, two complementary packages (yet again, among others!) in terms of ease of use and flexibility in the outcome are proposed: `{writexl}` and `{openxlsx}`.

As its name suggests, `{writexl}` is the extension of `{readxl}` dedicated to exporting tables to Excel through the `write_xlsx()` function. Its use is very simple as it only takes as inputs the table (or list of tables) to export to the file specified in the `path` parameter. 


```{r}
library(readxl)
library(writexl)
library(dplyr)

file_path <- file.path("data", "Sensory Profile.xlsx")

product_info <- read_excel(path  = file_path,
                           sheet = "Product Info",
                           range = "A1:D12",
                           col_names = TRUE)

#Basic data manipulation: writing a subset into a new xlsx
product_info %>% 
  filter(Protein %in% "High") %>% 
  write_xlsx(path = file.path("output", "High Protein Products.xlsx"),
             col_names = TRUE
             )

```



The export of tables using the `{writexl}` package is easy, yet simplistic as it does not allow formatting the tables (except for some minor possibilities for the header), nor does it allow exporting multiple tables within the same sheet. For more advanced exporting options, the use of **{openxlsx}** package is preferred as it allows more flexibility in structuring and formatting the Excel output.

With `{openxlsx}`, the procedure starts with creating a workbook object `wb` using the `createWorkbook()` function, to which we add worksheets through the `addWorksheet()` function. 
On a given worksheet, any table can be exported using `writeData()` or `writeDataTable()`, which controls where to write the table through the `startRow` and `startCol` options.
Through these different functions, many additional formatting procedure can be applied:  

- `createWorksheet()` allows:
  * show/hide grid lines using `gridLines`;
  * color the sheet using `tabColour`;
  * change the zoom on the sheet through `zoom`;
  * show/hide the tab using `visible`;
  * format the worksheet by specifying its size (`paperSize`) and orientation (`orientation`).
  
- `writeData()` and `writeDataTable()` allow:
  * controlling where to print the data using `startRow` and `startCol` (or alternatively `xy`: `xy = c("B",12)` prints the table starting in cell B12), hence allowing exporting multiple tables within the same sheet;
  * including the row names and column names through `rowNames` and `colNames`;
  * formatting the header using `headerStyle` (incl. color of the text and/or background, font, font size, etc.);
  * shaping the borders using predefined solutions through `borders`, or customizing them with `borderStyle` and `borderColour`;
  * adding a filter to the table using `withFilter`;
  * converts missing data to "#N/A" or any other string using `keepNA` and `na.string`.
  
- Additional formatting can be controlled using:
  * `options()` to pre-define number formatting, border colors and style that will be applied automatically to each table;
  * `modifyBaseFont()` to define the font name and font size;
  * `freezePane()` to freeze the first row and/or column of the table using `firstRow = TRUE` and `firstCol = TRUE`;
  * `createStyle()` to pre-define a style, or `addStyle()` to apply the styling to selected cells;
  * `setColWidths` to control column width;
  * `conditionalFormatting()` styling of cells when they meet pre-defined rules, as for instance to highlight significant p-values. 
  
When using `{openxlsx}`, we recommend to use the same procedure as for Word and PowerPoint: 

- Start with setting the default parameters that should be applied to each table;
- Create styles for text or table headers that you save in different elements, and that you apply where needed. 

In the following example, the sensory profiles are exported into a newly created sheet. 

To introduce conditional formatting with `{openxlsx}`, the sensory profiles are color coded as following: for each cell, the value is compared to the overall mean computed for that column and is colored in red (resp. blue) if it's higher (resp. lower) than the mean.
In practice, the color style is pre-defined in two parameters called `pos_style` (red) and `neg_style` (blue) using `createStyle()`. The decision whether `pos_style` or `neg_style` should be used is defined by the `rule` parameter from the `conditionalFormatting()`[^1] function.

[^1]: In `conditionalFormatting()`, you can specify to which `rows` and `cols` the formatting applies. In this example, `cols` takes `2` because the first column contains the row names.


```{r openxlsx}
library(openxlsx)

# Pre-define options to control the borders 
options("openxlsx.borderColour" = "#4F80BD")
options("openxlsx.borderStyle" = "thin")

# Automatically set Number formats to 3 values after the decimal
options("openxlsx.numFmt" = "0.000")

# Create a header style in which 
  # a blue background is used,
  # borders on top and on the bottom, 
  # the text is centered and is bold

headSty <- createStyle(fgFill = "#DCE6F1",
                       border = "TopBottom",
                       halign = "center",
                       textDecoration = "bold")

# Data preparation
senso_mean <- sensory %>% 
  group_by(Product) %>% 
  summarise(across(where(is.numeric),mean)) %>% 
  tibble::column_to_rownames(var = "Product")

overall_mean <- apply(senso_mean, 2, mean)

# Create workbook object
wb <- openxlsx::createWorkbook()

# Change the font to Calibri size 10
modifyBaseFont(wb,fontName = "Calibri", fontSize = 10)

# Add a new worksheet
addWorksheet(wb, sheetName = "Mean", gridLines = FALSE)

# Write table: note that 
writeDataTable(wb,
               sheet = "Mean",
               x = senso_mean, 
               startCol = 1,
               startRow = 1,
               colNames = TRUE, rowNames = TRUE, 
               tableStyle = "TableStyleLight9")

# Freezing the table
freezePane(wb, sheet = "Mean", firstRow = TRUE, firstCol = TRUE)

# Styles for conditional formatting
pos_style <- createStyle(fontColour = "firebrick3", bgFill = "mistyrose1")
neg_style <- createStyle(fontColour = "navy", bgFill = "lightsteelblue")

# Adding formatting to the second column
conditionalFormatting(wb,
                      sheet = "Mean", 
                      cols  = 2,
                      rows  = 1 + 1:nrow(senso_mean), 
                      rule  = paste0(">",overall_mean[2]),
                      style = pos_style)

conditionalFormatting(wb,
                      sheet = "Mean",
                      cols  = 2,
                      rows  = 1 + 1:nrow(senso_mean), 
                      rule  = paste0("<",overall_mean[2]),
                      style = neg_style)

setColWidths(wb, sheet = "Mean",
             cols = 1:(1+ncol(senso_mean)), widths = 12)

```


The file is created using `saveWorkbook()` by specifying the name of the workbook `wb` and its path through `file`. In case such workbook already exists, it can be overwritten using `overwrite`. 

TIPS: At any time, you can visualize your file using `openXL()`: This function opens within Excel the temporary file that is currently being built, hence allowing you to double-check that your code matches your expectations.

For more details on using **{openxlsx}** see <https://rdrr.io/cran/openxlsx/>.

## PowerPoint

Start with template
Explain slide master
How to adjust choices
Internal naming (relevant later)
Example
Title slides
Tables
Charts
Bullet points
Images
Layout discussion


```{r}
library(tidyverse)
library(officer)
library(flextable)
```

### PowerPoint Formatting

PowerPoint Slide Master

#### Importing the Template 

```{r 1.1 importing the template}
my_file <- file.path("input", "templates", "Tutorial Template.pptx") %>% 
                        read_pptx()

class(my_file) # checking if correct class

my_file %>%
  layout_summary()
```

#### Creating a PowerPoint Deck

```{r 1.2 creating a PPT deck}
pptx_obj <- read_pptx () # new empty file
pptx_obj <- pptx_obj %>% 
  add_slide(layout  = 'Title and Content', master = "Office Theme")

pptx_obj %>% print(target = "output/first_example.pptx")
```

We cannot load the themes of Office *ex nihilo* returns error
```{r themes}
pptx_obj <- read_pptx() # new empty file
pptx_obj <- pptx_obj %>% 
  add_slide(layout = "Title and Content", master = "Integral")
```

However, we can save an empty pptx with the desired theme and use it as a template
```{r themes1}
pptx_obj <- read_pptx(file.path("input", "templates", "integral.pptx"))
layout_summary(pptx_obj)
```

We can even load a template with more than one theme
```{r themes2}
pptx_obj <- read_pptx(file.path("input", "templates", "multmasters.pptx"))
layout_summary(pptx_obj)
```

#### Selection Pane

#### Key functions: 
`read_pptx(path)` `layout_summary(x)` `layout_properties(x)` 

`add_slide(x, layout, master)` `on_slide(x, index)` `slide_summary(x, index)`


#### Example Code
```{r 1.5 example code}
pptx_obj <- read_pptx() # new empty file

pptx_obj <- pptx_obj %>% # add slide
  add_slide(layout = "Title and Content", master = "Office Theme")

layout_summary(pptx_obj) # contains only basic layouts
layout_properties(pptx_obj) # additional detail

pptx_obj <- pptx_obj %>% 
  on_slide(index = 1) # set active slide
slide_summary(pptx_obj) # slide is empty
```

### Placeholders

#### Placeholders and Shapes

```{r 2.1 shapes1}
# Example 1
my_data <- c("My functions are:", "ph_with", "ph_location_type")
my_type <- "body"

pptx_obj <- read_pptx() # new empty file

pptx_obj <- pptx_obj %>% 
  add_slide(layout = "Title and Content", master = "Office Theme")

pptx_obj <- pptx_obj %>%
  ph_with(value = my_data, location = ph_location_type(type = my_type))

pptx_obj %>%
  print(target = "output/test2.1.pptx") 
```

```{r 2.2 shapes2}
# Example 2: 
my_data <- head(mtcars)[,1:4]
my_type <- "body"

pptx_obj <- read_pptx() # new empty file
pptx_obj <- pptx_obj %>% 
    add_slide(layout = "Title and Content", master = "Office Theme")

pptx_obj <- pptx_obj %>% 
  ph_with(value = my_data, location = ph_location_type(type = my_type)) 

pptx_obj %>%
  print(target = "output/test2.2.pptx") 
```

```{r 2.3 shapes3}
# Example 3
# We add a text box item in a custom position
# The same can be done for an image, logo, custom objects, etc.

my_data <- "My text"
my_type <- "body"

pptx_obj <- read_pptx() # new empty file
pptx_obj <- pptx_obj %>%
  add_slide(layout = "Title and Content", master = "Office Theme")

# ph_location is a subfunction which takes as argument 
# left/top/width/height, units are inches

pptx_obj <- pptx_obj %>%
  ph_with(value = my_data, location = ph_location(left = 2, top = 2, width = 3, height = 1))

pptx_obj %>%
  print(target = "output/test2.3.pptx") 
```

#### Key functions: ph_with()


### Text

#### Working with Text
Each new text item added to a PowerPoint via officer is a paragraph object
`fpar()` (“formatted paragraph”) creates this object

```{r 3.1 text1}
my_data <- fpar("My text")
my_type <- "body"

pptx_obj <- read_pptx() 
pptx_obj <- pptx_obj %>% 
  add_slide(layout = "Title and Content", master = "Office Theme")

# Add paragraph
pptx_obj <- pptx_obj %>% 
  ph_with(value = my_data, location = ph_location_type(type = my_type))

# Try to add a second paragraph
my_data2 <- fpar("My other text")
my_type <- "body"

pptx_obj <- pptx_obj %>% 
  ph_with(value = my_data2, location = ph_location_type(type = my_type) )

pptx_obj %>%
  print(target = "output/test3.1.pptx") 
# As we see, this code doesn’t produce bullet points as we might hope
```

`block_list()` allows us to wrap multiple paragraphs together

```{r 3.2 text2 blocklist}
my_data <- fpar("My text")
blank_line <- fpar("")
my_data2 <- fpar("My other text")

my_list <- block_list(my_data, blank_line, my_data2)

my_type <- "body"

pptx_obj <- read_pptx()
pptx_obj <- pptx_obj %>%
  add_slide(layout = "Title and Content", master = "Office Theme") %>% 
  ph_with(value = my_list, location = ph_location_type(type = my_type) )

pptx_obj %>%
  print(target = "output/test3.2.pptx") 
```

Use `ftext()` ("formatted text") to edit the text before pasting into paragraphs. `ftext()` requires a second argument called prop which contains the formatting properties.
```{r 3.3 text3}
my_prop <- fp_text(color = "red", font.size = 16)
my_text <- ftext("Hello", prop = my_prop)

my_par <- fpar(my_text) # formatted
blank_line <- fpar("")

my_par2 <- fpar("My other text") # unformatted
my_list <- block_list(my_par, blank_line, my_par2)

pptx_obj <- read_pptx()

pptx_obj <- pptx_obj %>%
  add_slide(layout = "Title and Content", master = "Office Theme") %>% 
  ph_with(value = my_list, location = ph_location_type(type = my_type))

pptx_obj %>%
  print(target = "output/test3.3.pptx") 
```

#### Key functions:
`fpar()` `ftext()` `fp_text()` `block_list()`

#### Example Code
```{r 3.4 example code}
my_list <- block_list(
  fpar(ftext("Hello", prop = fp_text(color = "red", font.size = 16))) ,
  fpar(ftext("World", prop = fp_text(color = "blue", font.size = 14)))
  )

# The hierarchy is: `block_list > fpar > ftext > fp_text`

pptx_obj <- read_pptx() %>%
  add_slide(layout = "Title and Content", master = "Office Theme") %>%
  ph_with(value = block_list(
    fpar(ftext("Hello", prop = fp_text(color = "red", font.size = 16))),
    fpar(ftext("World", prop = fp_text(color = "blue", font.size = 14)))
    ),
    ph_location_type(type = "body")
  ) %>%
  print(target = "output/test3.4.pptx") 
```

### Tables
#### Basic Code
```{r 4.1 basic code}
ft_data <- senso_mean %>%
  dplyr::select(Salty, Sweet, Sour, Bitter) %>% 
  tibble::rownames_to_column() %>% 
  rename(Product = rowname) %>% 
  mutate(across(Salty:Bitter, as.numeric)) %>% 
  bind_rows(summarise(.,
                      across(where(is.numeric), mean),
                      across(where(is.character), ~"Average"))) %>% 
  mutate(across(where(is.numeric), round, 2)) 

pptx_obj <- read_pptx() %>%
  add_slide(layout = "Title and Content", master = "Office Theme") %>%
  ph_with(value = ft_data, location = ph_location_type(type = "body")) %>%
  print(target = "output/test4.1.pptx") 
```

#### Introduction to flextable

With `{flextable}`, the procedure starts with creating a flextable object `ft_table` using the `flextable()` function.
Flextable objects are compatible with `{officer}` and therefore are our primary tool for table formatting.
Through different functions, many custom formatting procedure can be applied:  

#### key functions:
 `align()` `bold()` `font()` `color()` `bg()` `height()` & `width()` 
 `border_outer()` & `border_inner()` & `border_inner_h()` `border_inner_v()` `autofit()`

Additional function to learn: `merge()`, `compose()` & `as_chunk()`, `style()`

```{r flextable2}
library(flextable)

# Create a flextable object
ft_table <- ft_data %>% 
  flextable()

# Flextable objects can be found in the Viewer tab of Rstudio
print(ft_table)

```

#### Formatting examples

```{r 4.4 demonstration}
ft_table <- ft_table %>%
  autofit() %>% # column width
  # alignment of header: we use part argument
  align(align = "center", part = "header") %>%
  # alignment of content: we can use part = "body" or specify exact lines 
  align(i = 1:nrow(ft_data), j = 1:ncol(ft_data), align = "center") 

print(ft_table)

```

Set font names, sizes and colors
```{r 4.5 demonstration cont.1}
ft_table <- ft_table %>% 
  # main formatting
  fontsize(size = 11) %>%
  font(fontname = "Calibri") %>% # since no i or j are input, change is for all data
  font(fontname = "Roboto", part = "header") %>% #different font for header
  color(color = "white", part = "header") %>%
  bold(part = "header") %>%
  # format last row
  bold(i = nrow(ft_data), j = 1:ncol(ft_data)) %>% #
  italic(i = nrow(ft_data), j = ~Product + Salty + Sweet + Sour + Bitter) %>% # using ~ notation
  color(i =  nrow(ft_data), j = ~Sour, color = "red") %>%
  color(i =  nrow(ft_data), j = ~Sweet, color = "orange") %>%
  # background colors
  bg(bg = "#324C63", part = "header") %>% # a custom background for the header
  bg(i = 1:nrow(ft_data), bg = "#EDEDED") # a custom background for some cells

print(ft_table)
```

Set borders and adjust cells heights and widths
```{r 4.6 demonstration cont.2}
#BORDERS
# For borders we need to use nested functions (similar to fpar>ftext>fp_text)
# fp_border() is the second level function we will use to specify border"s characteristics
# as argument it takes color, style, and width
my_border <- officer::fp_border(color = "black", style = "solid", width = 1)

# We use this second level function inside various main border functions
# border_outer(), border_inner(), border_inner_h(), border_inner_v()
ft_table <- ft_table %>%
  border_outer(part = "all", border = my_border) %>% # using predefined border
  border_inner(part = "body", border = officer::fp_border(style = "dashed")) %>% 

  # all measurements are in inches
  width(j = 1, width = 1.2) %>% # column 1 wider
  height(i = 12, height = 1) # last row's height

ft_table
```

#### Demonstration Output

#### Add flextable object to a powerpoint slide

```{r 4.7 original example}
# Add table to slide
pptx_obj <- read_pptx() %>%
  add_slide(layout = "Title and Content", master = "Office Theme") %>%
  ph_with(value = ft_table, ph_location(left = 2, top = 2, width = 4)) %>% 
  print(target = "output/test4.7.pptx") 
# positions are fixed. We can input exact positions to center the table

```


### Charts

#### Adding charts as images

```{r 5.1 rvg package}
# We have preloaded a function to plot the chart.
# the function is using ggplot2 as plotting library
chart_to_plot <- sample_data_list[['Sample 1']] %>%
  make_jar_chart() # code to create a ggplot2 item, we will skip the contents
print(chart_to_plot) # see in Plots Window
```

#### rvg Example

```{r 5.2 rvg example}
# the output is a ggplot2 object

# To add this object as a rvg object on a slide, we will use the ph_with_vg
# ph_with_vg replaces the ph_with for a rvg object
# ph_with_vg_at allows to input a precise position for a chart, using the top/left we know already
# all units are in inches
# argument code requires print(chart), argument type is a specific place on slide ("body" or other)

## IMPORTANT: ph_with_vg is is deprecated.
#old syntaxis ph_with_vg(code = print(chart_to_plot), type = "body") 

pptx_obj <- read_pptx() %>%
  add_slide(layout = "Title and Content", master = "Office Theme") %>%
  ph_with(value = dml(ggobj = chart_to_plot), location  =  ph_location_type(type = 'body')) %>%
  print(target = "output/5.2.rvg1.pptx") 

## IMPORTANT: ph_with_vg_at is is deprecated.
# old syntaxis ph_with_vg_at(code = print(chart_to_plot), left = 1, top = 1, width = 8, height = 6) 

pptx_obj <- read_pptx() %>%
  add_slide(layout = "Title and Content", master = "Office Theme") %>%
  ph_with(value = dml(ggobj = chart_to_plot), location =  ph_location(left = 1, top = 1, width = 8, height = 6))  %>%
  print(target = "output/5.2.rvg2.pptx") 

# all items on the chart inside the pptx are now editable, just click on any and see 
# the Shape Format tab in PowerPoint
```

### mschart Package

```{r 5.3 mschart example}
# sample dataframe
mydata <- sample_data_list[['Sample 1']] %>%
  group_by(Variable,Response) %>% count()

# syntaxis is similar to ggplot2"s aes() with x,y,group
my_barchart <- ms_barchart(data = mydata, x = "Variable", y = "n", group = "Response")

# to add the object to a powerpoint slide we can use the officer"s native ph_with
pptx_obj <- read_pptx() %>%
  add_slide(layout = "Title and Content", master = "Office Theme") %>%
  ph_with(value = my_barchart, location = ph_location_type(type = "body")) %>%
  print(target = "output/5.3.msoffice.pptx") 

# if we would open a rvg slide and an ms office slide and click on the slide
# for the rvg slide the only menu that appear is shape format
# While for the msoffice chart we have now the chart design option with all msoffice functionalities
# by using chart_settings() functions one can customise in R the charts
```

## Word

### Word documents
Formats share similarities

`body_add_*()`

```{r 6.1 word doc1}
my_doc <- read_docx() %>% 
  body_add_par(value = "My Text", style = "Normal") %>%
  body_add_par(value = "Other Text", style = "Normal") %>% 
  body_add_par(value = "Conclusion", style = "Normal") %>% 
  print(target = 'output/6.1.mydoc.docx')
```

`body_add_break()`

```{r 6.2 word doc2}
my_doc <- read_docx() %>% 
  body_add_par(value = "My Text", style = "Normal") %>%
  body_add_break() %>% 
  body_add_par(value = "Conclusion", style = "Normal") %>% 
  print(target = 'output/6.2.mydoc.docx')
```

```{r 6.3 word doc3}
my_format <- fp_text(font.family = 'Calibri', font.size = 14, bold = TRUE, color = 'red')
my_text <- ftext('My dataset is:', my_format)
my_par <- fpar(my_text)

doc <- read_docx() %>% 
  body_add_par(value = "Table of content", style = "heading 1") %>% 
  body_add_par(value = "", style = "Normal") %>% 
  body_add_fpar(my_par, style = "Normal") %>% #formatted paragraph function
  body_add_par(value = "", style = "Normal") %>% 
  body_add_table(value = head(mtcars)[, 1:4], style = "table_template" ) %>% 
  print(target = 'output/6.3.mydoc.docx')

read_docx() %>% styles_info() 
```



