```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```
# Data Collection {#data-collection}

## Background
## Other details
## Importing Data to R

To analyze data, one need *data*. If this data is already available in R, then the analysis can be performed directly. However, in much cases, the data is stored outside the R environment, and needs to be imported.

In practice, the data might be stored in as many format as one can imagine, whether it ends up being a fairly common solution (.txt file, .csv file, or .xls/.xlsx file), or software specific (e.g. Stata, SPSS, etc.).
Since it is very common to store the data in Excel spreadsheets (.xlsx) due to its simplicity, the emphasis is on this solution. Fortunately, most generalities presented for Excel files also apply to other formats through `base::read.table()` for .txt files,  `base::read.csv()` and `base::read.csv2()` for .csv files, or through the `{read}` package (which is part of the `{tidyverse}`).

For other (less common) formats, the reader can find packages that would allow importing their files into R. Particular interest can be given to the package `{rio}` (*rio* stands for *R* *I*nput and *O*utput) which provides an easy solution that 1. can handle a large variety of files, 2. can actually guess the type of file it is, and 3. provides tools to import, export, and convert almost any type of data format, including .csv, .xls and .xlsx, or data from other statistical software such as SAS (.sas7bdat and .xpt), SPSS (.sav and .por), or Stata (.dta). As an alternative, the package `{foreign}` provides functions that allow importing data stored from other statistical software (incl. Minitab, S, SAS, Stata, SPSS, etc.)

Although Excel is most likely one of the most popular way of storing data, there are no `{base}` functions that allow importing such files easily. Fortunately, many packages have been developed in that purpose, including `{XLConnect}`, `{xlsx}`, `{gdata}`, and `{readxl}`. Due to its convenience and speed of execution, we will be using `{readxl}` here.

### Importing Structured Excel File

First, let's import the *Sensory Profile.xlsx* workbook using the `readxl::read_xlsx()` file, by informing as parameter the location of the file (informed in `file_path` using the package `{here}`) and the `sheet` we want to read from.

This file is called _structured_ as all the relevant information is already stored in the same sheet in a structured way. In other words, no decoding is required here, and there are no 'unexpected' rows or columns (e.g. empty lines, or lines with additional information regarding the data but that is not data):

 - The first row within the *Data* sheet of *Sensory Profile.xlsx* contains the headers,  
 - From the second row onwards, only data is being stored.
 
Since this data will be used for some analyses, it is assigned data to an R object called `sensory`.

```{r import_sensory, echo=FALSE}
library(here)
file_path <- here("data","Sensory Profile.xlsx") 

library(readxl)
sensory <- read_xlsx(file_path, sheet="Data")

```

To ensure that the importation went well, we print `sensory` to see how it looks like. Since `{readxl}` has been developed by Hadley Wickham and colleagues, its functions follow the `{tidyverse}` principles and the dataset thus imported is a `tibble`. Let's take advantage of the printing properties of a `tibble` to evaluate `sensory`:

```{r}
sensory
```

`sensory` is a tibble with 99 rows and 35 columns that includes the `Judge` information (first column, defined as character), the `Product` information (second and third columns, defined as character), and the sensory attributes (fourth column onward, defined as numerical or `dbl`).

### Importing Unstructured Excel File

In some cases, the dataset is not so well organized/structured, and may need to be *decoded*. This is the case for the workbook entitled *TFEQ.xlsx*. For this file:

 - The variables' name have been coded and their corresponding names (together with some other valuable information we will be using in the next chapter) are stored in a different sheet entitled _Variables_;
 - The different levels of each variable (including their code and corresponding names) are stored in another sheet entitled _Levels_.

To import and decode this dataset, multiple steps are required:

 - Import the variables' name only;
 - Import the information regarding the levels;
 - Import the dataset without the first line of header, but by providing the correct names obtained in the first step;
 - Decode each question (when needed) by replacing the numerical code by their corresponding labels.

Let's start with importing the variables' names from *TFEQ.xlsx* (sheet *Variables*)

```{r import_TFEQ_labels1, echo=FALSE}
file_path <- here("data","TFEQ.xlsx") 

var_names <- read_xlsx(file_path, sheet="Variables")
var_names

```

In a similar way, let's import the information related to the levels of each variable, stored in the *Levels* sheet. 
A deeper look at the *Levels* sheet shows that only the coded names of the variables are available. In order to include the final names, `var_names` is joined (using `inner_join`).

```{r import_TFEQ_labels2}
library(tidyverse)
var_labels <- read_xlsx(file_path, sheet="Levels") %>% 
  inner_join(dplyr::select(var_names, Code, Name), by=c(Question="Code"))

var_labels

```

**Note**: In some cases, this information is directly available in the dataset as sub-header: A solution is then to import the first rows of the dataset that contain this information using the parameter `n_max` from `readxl::read_xlsx``. For each variable (when information is available), store that information as a list of tables that contains the code and their corresponding label.

<!-- **THIS SECTION BELOW MIGHT NEED TO GET PASSED ON TO THE EXERCISE**

Since most likely this system of coding follow a fixed pattern, we strongly recommend the use of `{tidytext}` and its function `unnest_tokens()`.
For example, let's imagine that the our information is structured as *code1=label1,code2=label2,...* (e.g. *0=No,1=Yes*). In such case, first use `unnest_tokens()` to split this string by ','. This creates a tibble with as many rows as there are *code=label* and one column. Next, split this column into two columns using `separate()` and `sep="="`.

**(PREVIOUS PART) TO BE GIVEN AS AN EXAMPLE/EXERCISE** -->

Finally, the dataset (*Data*) is imported by substituting the coded names with their corresponding names.
This process can be done by skipping reading the first row of the dataset that contains the coded header (`skip=1`), and by passing `Var_names` as header or column names (after ensuring that the names' sequence perfectly match across the two tables!).

```{r, import_TFEQ_data3, echo=FALSE}
TFEQ_data <- read_xlsx(file_path, sheet="Data", col_names=var_names$Name, skip=1)

```

The data has now the proper header, however each variable is still coded numerically. The steps to convert the numerical values with their corresponding labels is shown in Section \@ref(data-prep).
