```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# Machine Learning {#machine_learning}

## Overview

## Key Topics

### Model Validation
### Unsupervised learning
#### Cluster analysis
#### Factor analysis
#### Principle components analysis
#### t-SNE
### Semisupervised learning
#### PLS regression
#### Cluster Characterization
### Supervised learning
#### Regression
#### K-nearest neighbors
#### Decision trees
#### Black boxes
##### Random forests
##### SVMs
##### Neural networks
##### Computer vision
### Interpretability 
#### LIME
#### DALEX
#### IML

## Common Applications
### Predicting sensory profiles from instrumental data
### Predicting consumer response from sensory profiles
### Characterizing consumer clusters

## Code Examples

### Data Prep

```{r data-prep}

data <- readr::read_rds('data/masked_data.rds')
nrows <- max(summary(data$Class)) * 2

data_over <- ROSE::ROSE(Class ~ .,
                        data = data %>% 
                          mutate(across(starts_with('D'), factor, levels = c(0, 1))),
                        N = nrows, seed = 1)$data

readr::write_rds(data_over, 'data/data_classification.rds')

readxl::read_excel('data/data_regression.xlsx') %>%
  select(-`...1`, -judge, -product, -(steak:V64), -`qtt.drink.(%)`) %>%
  rename(socio_professional = `socio-professional`) %>%
  readr::write_rds('data/data_regression.rds')


```


### Classification Code

```{r classification-code}

library(tidyverse)
library(tidymodels)


# Load data ---------------------------------------------------------------

data <- read_rds('data/data_classification.rds')

# Inspect the data --------------------------------------------------------

summary(data)

data <- data %>% select(-ID)

skimr::skim(data)

data %>%
  mutate(across(starts_with('D'), factor, levels = c(0, 1))) %>%
  GGally::ggpairs(aes(fill = Class))



# Split data for models ---------------------------------------------------

# Set test set aside
train_test_split <- initial_split(data)
train_test_split

train_set <- training(train_test_split)
test_set <- testing(train_test_split)

# Split set fot cross-validation
resampling <- vfold_cv(train_set, 10)
resampling


# Fit MARS model ----------------------------------------------------------

usemodels::use_earth(
  Class ~ .,
  data = train_set
  )

earth_recipe <- 
  recipe(formula = Class ~ ., data = train_set) %>% 
  step_novel(all_nominal(), -all_outcomes()) %>% 
  step_dummy(all_nominal(), -all_outcomes()) %>% 
  step_zv(all_predictors()) 

earth_spec <- 
  mars(
    num_terms = tune(),
    prod_degree = tune(),
    prune_method = "none"
    ) %>% 
  set_mode("classification") %>% 
  set_engine("earth") 

earth_workflow <- 
  workflow() %>% 
  add_recipe(earth_recipe) %>% 
  add_model(earth_spec) 

earth_grid <- tidyr::crossing(num_terms = 2 * (1:6), prod_degree = 1:2) 
earth_grid

earth_tune <- 
  tune_grid(
    earth_workflow, 
    resamples = resampling, 
    # Save predictions for further steps
    control = control_grid(save_pred = TRUE, verbose = TRUE),
    # Test parameters on a grid defined above
    grid = earth_grid
  ) 


# Check model performance -------------------------------------------------

earth_tune %>% show_best(n = 10)
earth_tune %>% autoplot()

earth_predictions <- earth_tune %>%
  collect_predictions(parameters = select_best(., 'roc_auc')) %>%
  mutate(model = "MARS")

earth_predictions %>%
  roc_curve(Class, .pred_A) %>%
  autoplot()

earth_predictions %>%
  lift_curve(Class, .pred_A) %>%
  autoplot()

earth_predictions %>%
  pr_curve(Class, .pred_A) %>%
  autoplot()

earth_predictions %>%
  conf_mat(Class, .pred_class) %>%
  autoplot()


# Fit decision tree -------------------------------------------------------

tree_recipe <- 
  recipe(formula = Class ~ ., data = train_set) %>% 
  step_novel(all_nominal(), -all_outcomes()) %>% 
  step_zv(all_predictors()) 

tree_spec <- 
  decision_tree(
    cost_complexity = tune(),
    tree_depth = tune(),
    min_n = tune()
    ) %>% 
  set_mode("classification") %>% 
  set_engine("rpart") 

tree_workflow <- 
  workflow() %>% 
  add_recipe(tree_recipe) %>% 
  add_model(tree_spec) 

tree_tune <- 
  tune_grid(
    tree_workflow, 
    resamples = resampling, 
    # Save predictions for further steps
    control = control_grid(save_pred = TRUE, verbose = TRUE),
    # Test 20 random combinations of parameters
    grid = 20
  ) 

# Check model performance -------------------------------------------------

tree_tune %>% show_best(n = 10)
tree_tune %>% autoplot()

tree_predictions <- tree_tune %>%
  collect_predictions(parameters = select_best(., 'roc_auc')) %>%
  mutate(model = "Decision Tree")

tree_predictions %>%
  bind_rows(earth_predictions) %>%
  group_by(model) %>%
  roc_curve(Class, .pred_A) %>%
  autoplot()

tree_predictions %>%
  bind_rows(earth_predictions) %>%
  group_by(model) %>%
  lift_curve(Class, .pred_A) %>%
  autoplot()

tree_predictions %>%
  bind_rows(earth_predictions) %>%
  group_by(model) %>%  pr_curve(Class, .pred_A) %>%
  autoplot()

tree_predictions %>%
  conf_mat(Class, .pred_class) %>%
  autoplot()


# Let's go with MARS model ------------------------------------------------

final_fit <- earth_workflow %>%
  finalize_workflow(select_best(earth_tune, 'roc_auc')) %>%
  last_fit(train_test_split)

final_fit %>% collect_metrics()

final_fit %>%
  collect_predictions() %>%
  roc_curve(Class, .pred_A) %>%
  autoplot()

final_model <- final_fit %>%
  pluck(".workflow", 1) %>%
  fit(data)

final_model %>%
  pull_workflow_fit() %>% 
  vip::vip()

final_model %>%
  pull_workflow_fit() %>%
  pluck("fit") %>%
  summary

write_rds(final_model, 'classification_model.rds')


# Predict something -------------------------------------------------------

model <- read_rds('classification_model.rds')

new_observation <- tibble(
  N1 = 1.8,
  D1 = factor(0),
  D2 = factor(0),
  D3 = factor(1),
  D4 = factor(0),
  D5 = factor(1),
  D6 = factor(0),
  D7 = factor(1),
  D8 = factor(1),
  D9 = factor(1),
  D10 = factor(1),
  D11 = factor(0)
)

predict(model, new_observation, type = "class")
predict(model, new_observation, type = "prob")





```



### Regression Code

```{r regression-code}

library(tidyverse)
library(tidymodels)


# Load data ---------------------------------------------------------------

data <- read_rds('data/data_regression.rds')
glimpse(data)

# Inspect the data --------------------------------------------------------

summary(data)

skimr::skim(data)

# Split data for models ---------------------------------------------------

# Set test set aside
train_test_split <- initial_split(data)
train_test_split

train_set <- training(train_test_split)
test_set <- testing(train_test_split)

# Split set fot cross-validation
resampling <- vfold_cv(train_set, 10)
resampling


# Fit glmnet model ----------------------------------------------------------

usemodels::use_glmnet(
  Liking ~ .,
  data = train_set
)

glmnet_recipe <- 
  recipe(formula = Liking ~ ., data = train_set) %>% 
  step_novel(all_nominal(), -all_outcomes()) %>% 
  step_dummy(all_nominal(), -all_outcomes()) %>% 
  step_zv(all_predictors()) %>% 
  step_normalize(all_predictors(), -all_nominal())

glmnet_spec <- 
  linear_reg(penalty = tune(), mixture = tune()) %>% 
  set_mode("regression") %>% 
  set_engine("glmnet") 

glmnet_workflow <- 
  workflow() %>% 
  add_recipe(glmnet_recipe) %>% 
  add_model(glmnet_spec) 

glmnet_grid <- tidyr::crossing(penalty = 10^seq(-6, -1, length.out = 20), 
                               mixture = c(0.05, 0.2, 0.4, 0.6, 0.8, 1)) 

glmnet_tune <- 
  tune_grid(
    glmnet_workflow, 
    resamples = resampling, 
    # Save predictions for further steps
    control = control_grid(save_pred = TRUE, verbose = TRUE),
    # Test parameters on a grid defined above
    grid = glmnet_grid
    ) 

# Check model performance -------------------------------------------------

glmnet_tune %>% show_best(n = 10)
glmnet_tune %>% autoplot()

glmnet_predictions <- glmnet_tune %>%
  collect_predictions(parameters = select_best(., 'rmse')) %>%
  mutate(model = "GLMNet",
         .resid = Liking - .pred)

glmnet_predictions %>%
  ggplot(aes(sample = .resid)) +
  geom_qq() +
  geom_qq_line()

glmnet_predictions %>%
  ggplot(aes(.pred, Liking)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0)

glmnet_predictions %>%
  ggplot(aes(.pred, .resid)) +
  geom_point() +
  geom_hline(yintercept = 0)

ggplot(glmnet_predictions, aes(x = .resid)) + 
  geom_histogram(aes(y =..density..), fill = 'white', color = 'black') +
  stat_function(fun = dnorm,
                args = list(mean = mean(glmnet_predictions$.resid), 
                            sd = sd(glmnet_predictions$.resid)),
                size = 1)

# Fit random forest -------------------------------------------------------

rf_recipe <- 
  recipe(formula = Liking ~ ., data = train_set) %>% 
  step_novel(all_nominal(), -all_outcomes()) %>% 
  step_zv(all_predictors()) 

rf_spec <- 
  rand_forest(
    mtry = tune(), 
    min_n = tune(),
    trees = 50
    ) %>% 
  set_mode("regression") %>% 
  set_engine("ranger", importance = "impurity") 

rf_workflow <- 
  workflow() %>% 
  add_recipe(rf_recipe) %>% 
  add_model(rf_spec) 

rf_tune <- 
  tune_grid(
    rf_workflow, 
    resamples = resampling, 
    # Save predictions for further steps
    control = control_grid(save_pred = TRUE, verbose = TRUE),
    # Test 20 random combinations of parameters
    grid = 20
  ) 

# Check model performance -------------------------------------------------

rf_tune %>% show_best(n = 10)
rf_tune %>% autoplot()

rf_predictions <- rf_tune %>%
  collect_predictions(parameters = select_best(., 'rmse')) %>%
  mutate(model = "Random Forest",
         .resid = Liking - .pred)

rf_predictions %>%
  bind_rows(glmnet_predictions) %>%
  ggplot(aes(sample = .resid)) +
  geom_qq() +
  geom_qq_line() +
  facet_wrap(~model)

rf_predictions %>%
  bind_rows(glmnet_predictions) %>%
  ggplot(aes(.pred, Liking)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0) +
  facet_wrap(~model)

rf_predictions %>%
  bind_rows(glmnet_predictions) %>%
  ggplot(aes(.pred, .resid)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  facet_wrap(~model)

rf_predictions %>%
  ggplot(aes(x = .resid)) + 
  geom_histogram(aes(y =..density..), fill = 'white', color = 'black') +
  stat_function(fun = dnorm,
                args = list(mean = mean(rf_predictions$.resid), 
                            sd = sd(rf_predictions$.resid)),
                size = 1)

# Let's go with rf model ------------------------------------------------

final_fit <- glmnet_workflow %>%
  finalize_workflow(select_best(glmnet_tune, 'rmse')) %>%
  last_fit(train_test_split)

final_fit <- rf_workflow %>%
  finalize_workflow(select_best(rf_tune, 'rmse')) %>%
  last_fit(train_test_split)

final_fit %>% collect_metrics()

final_fit %>%
  collect_predictions() %>%
  mutate(.resid = Liking - .pred) %>%
  ggplot(aes(sample = .resid)) +
  geom_qq() +
  geom_qq_line()

final_model <-  final_fit %>%
  pluck(".workflow", 1) %>%
  fit(data)

final_model %>%
  pull_workflow_fit() %>%
  vip::vip()

# final_model %>%
#   broom::tidy() %>%
#   filter(estimate != 0)

write_rds(final_model, 'regression_model.rds')

# Predict something -------------------------------------------------------

model <- read_rds('regression_model.rds')

new_observations <- data[1:2,]
new_observations

predict(model, new_observations)




```


