```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```
# Getting Started in R {#start-R}

## R

R is an open-source programming language and software environment
First released in 1995, R is an open-source implementation of S
R was developed by Ross Ihaka and Robert Gentleman
The name “R” is partly a play on Ihaka’s and Gentleman’s first names
R is a scripting language (not a compiled language)
Lines of R code run (mostly) in order
R is currently the 7th most popular programming language in the world

### Why Learn a Programming Language?

Control
Speed
Reduced errors
Increased capability
Continuous improvement
Improved collaboration
Reproducible results

### Why R?

R originated as a statistical computing language
It has a culture germane to sensory science
R is well-supported with an active community
Extensive online help is available
Many books, courses, and other educational material exist
The universe of available packages is vast
R excels at data manipulation and results reporting
R has more specialized tools for sensory analysis than other programming language

## Why R?

For sensory and consumer scientists, we recommend the R ecosystem of tools for three main reasons.  The first reason is cultural - R has from its inception been oriented more towards statistics than to computer science, making the feeling of programming in R more natural (in our experience) for sensory and consumer scientists than programming in Python.  This opinion of experience is not to say that a sensory and consumer scientist shouldn't learn Python if they are so inclined, or even that Python tools aren't sometimes superior to R tools (in fact, they sometimes are).  This latter point leads to our second reason, which is that R tools are typically better suited to sensory and consumer science than are Python tools.  Even when Python tools are superior, the R tools are still sufficient for sensory and consumer science purposes, plus there are many custom packages such as SensR, SensoMineR, and FactorMineR that have been specifically developed for sensory and consumer science.  Finally, the recent work by the RStudio company, and especially the exceptional work of Hadley Wickham, has lead to a very low barrier to entry for programming within R together with exceptional tools for data manipulation.


### Steps to Install R

The first step in this journey is to install R.  For this, visit [The R Project for Statistical Computing](https://www.r-project.org/).  From there, follow the download instructions to install R for your particular platform.

https://cran.r-project.org/bin/windows/base/
Download the latest version of R
Install R with default options
You will almost certainly be running 64-bit R
Note: If you are running R 4.0 or higher, you might need to install Rtools:
https://cran.r-project.org/bin/windows/Rtools/


## RStudio



### Steps to Install RStudio


Next you need to install RStudio, which is our recommended integrated development environment (IDE) for developing R code.  To do so, visit the [RStudio desktop download page](https://rstudio.com/products/rstudio/download/) and follow the installation instructions.

Once you have installed R and RStudio, you should be able to open RStudio and enter the following into the Console to receive the number "3" as your output:

```{r}

x <- 1
y <- 2

x + y

```


Some recommendations upon installing RStudio:

  - Change the color scheme to dark.
  - Put the console on the right.
  

https://www.rstudio.com/products/rstudio/download/#download
Download and install the latest (almost certainly 64-bit) version of RStudio with default options
Adjustments:
Uncheck “Restore .RData into workspace at startup
Select “Never” for “Save workspace to .RData on exit”
Change color scheme to dark (e.g. “Idle Fingers”)
Put console on right

### Create a Local Project

Always work in an RStudio project
Projects keep your files (and activity) organized
Projects help manage your file path (so your computer can find things)
Projects allow for more advanced capabilities later (like GitHub or renv)
We cover the use of GitHub in a future webinar
For now we create projects locally

### Install and Load Packages

As you use R, you will want to make use of the many packages others (and perhaps you) have written
Essential packages (or collections):
tidyverse, readxl
Custom Microsoft office document creation
officer, flextable, rvg, openxlsx, extrafont, extrafontdb
Sensory specific packages
sensR , SensoMineR, FactoMineR, factoextra
There are many more, for statistical tests of all varieties, to multivariate analysis, to machine learning, to text analysis, etc.

You only need to install each package once per R version
To install a package, you can:
Type install.packages("[package name]")
Use the RStudio dropdown
In addition, if a script loads package that are not installed, RStudio will prompt you to install the package
Notes: 
If you do not have write access on your computer, you might need IT help to install packages
You might need to safelist various R related tools and sites 

### Run Sample Code

Like any language, R is best learned first through example then through study
We start with a series of examples to illustrate basic principles
For this example, we analyze a series of Tetrad tests

Suppose you have 15 out of 44 correct in a Tetrad test
Using sensR, it’s easy to analyze these data:

```{r first-example}

library(sensR)

num_correct <- 15  
num_total <- 44  
  
discrim_res <- discrim(num_correct, num_total, method = "tetrad")  
  
print(discrim_res)  


```

## Git and GitHub

Git is a version control system that allows you to revert to earlier versions of your code, if necessary.  GitHub is service that allows for online backups of your code and which facilitates collaboration between team members.  We highly recommend that you integrate both Git and GitHub into your data scientific workflow.  For a full review of Git and GitHub from an R programming perspective, we recommend [Happy Git with R](https://happygitwithr.com/) by Jenny Bryant.  In what follows, we simply provide the minimum information needed to get you up and running with Git and GitHub.  Also, for an insightful discussion of the need for version control, please see [Cite bryan2018excuse].

### Git
  
  - Install Git
    * Windows
    * macOS
  - Register with RStudio

  
### GitHub

  - Create a GitHub account
  - Register with RStudio

## RAW MATERIAL
  
### Principles
### Tools
#### GitHub
#### R scripts
#### RMarkdown
#### Shiny
### Documentation
### Version control
### Online repositories for team collaboration
### Building a code base
#### Internal functions
#### Packages

  
