---
title: "My first Markdown document"
author: "John, Julien, Thierry"
date: "11/3/2021"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

```

# Introduction to R Markdown

This is my first introduction to R Markdown and the `{rmarkdown}` package, which is an **awesome** package.

# Analysis 

## Outline of this section

In this section, I am going to run to do two things:

1. Run some code
2. Export some results
  - A table generated by R
  - Some results within the text
  - A Figure with its caption. 
  
## Analysis

Let's take a simple example using the `cars` data set, in which we *regress* distance on speed. 

```{r}

fit = lm(dist ~ speed, data=cars)
b = coef(fit)

```

The slope of the regression is `r b[2]`.

The results of the regression a shown in two different ways, first directly from R, and second after formatting the table using the `knitr::kable()` function:

```{r}

table <- broom::tidy(fit)

table

knitr::kable(table)

```

Let's also visualize the results graphically:

```{r first-plot, fig.cap="My first Plot"}
plot(cars)
abline(fit)
```

The results are showed in Figure \@ref(fig: first-plot).